---
title: "test"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(echo = TRUE, eval=FALSE)

```

# llll

```{r import, echo=FALSE}
library(illuminaHumanv4.db)
library(ggpubr)
library(GEOquery)
library(limma)
library(gplots)
library(dplyr)
library(tidyr)



```

## Importing the Data

```{r getting_experiment,  echo=FALSE}
my_id <- "GSE17992"
gse <- getGEO(my_id, GSEMatrix =TRUE)
```

Getting the samples from platform GPL570 from the experiment (in the bioproject)
```{r get_samples, echo=FALSE}
if (length(gse) > 1) idx <- grep("GPL4133", attr(gse, "names")) else idx <- 1
gset1 <- gse[[idx]]

gset = gset1

```


Seeing the metadata
```{r metadata}
pData(gset)
fData(gset)
fvarLabels(gset)
```
```{r}
df = fData(gset)
filtered_df <- subset(df, GB_ACC == "NM_021130")

```

Seeing the headers of the metadata
```{r headers}
sample_data = pData(gset)
colnames(sample_data)

```


```{r clean_up}
sample_data$age = sample_data$`agent:ch1`


```



Getting the microarray ID's data (like the gene name) but this is RNA seq. I have the counts with gene symbols

```{r sample_metadata}
fvarLabels(gset) <- make.names(fvarLabels(gset))
features <- fData(gset)
features
```

Here, I will need a function to take a column and gives me a vector, 0 is it is young, 1 if it is middle age and 2 if it is old
```{r age_category_function}
map_age <- function(age_category) {
  
  age_vector <- ifelse(age_category == "control", 0, 1)
  #age_vector <- ifelse(age_category < 35, 0, ifelse(age_category >= 65, 2, 1))

  return (paste(age_vector, collapse = ""))
}

```

If we use it, we will get the mapped data

```{r }
# coding subtypes, 0 = YOUNG, 1 = MIDDLE AGE, 2 = OLD
gsms <- map_age(sample_data$age)
gsms #<- "111111111111000000000000"
sml <- strsplit(gsms, split="")[[1]]
```
And this is to put the labels when we don't have them, but in this case we do
```{r mapping}

gs <- factor(sml)
groups <- make.names(c("control", "trearment"))
levels(gs) <- groups
gset$group <- gs
```
But since we did when we separated the title, we go with


## Check normalisation and scales used

Log2 scale transformation required

Check for normalisation by looking at boxplot of expression levels for each sample


```{r }
ex <- exprs(gset) # expression data, what form is this?
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
exprs(gset) <- log2(ex) } # i think this code is trying to determine whether the data is in log2 format yet
```

And here we see how are the samples before any selection
```{r boxplot1}

ord <- order(gs)  # order samples by group

palette <- palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste (my_id, "/", annotation(gset), sep ="")
boxplot(log2(ex[,ord]), boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")
```


```{r boxplot2}
boxplot(log2(ex),range=0,ylab="log2 intensity")
```


## Probe filtering

```{r }
# expression value distribution
par(mar=c(4,4,2,1))
title <- paste ("Normalized Expression Density", sep ="")
plotDensities(ex[,ord], group=gs, main=title, legend ="topright")
```

```{r}
# Probe filtering
medians <- rowMedians(ex)
man_threshold <- 4 

hist(medians, 150, col = "cornsilk1", freq = FALSE, 
main = "Histogram of the median intensities", 
border = "antiquewhite4",
xlab = "Median intensities")
```

```{r}
meta <- pData(gset)
meta$group
```

```{r}
no_of_samples <- table(meta$group)
sample_cutoff <- min(no_of_samples)

idx_man_threshold <- apply(Biobase::exprs(gset), 1, function(x){ sum(x > man_threshold) >= sample_cutoff})

table(idx_man_threshold)
```


```{r}
batches <- as.factor(gset$group)

col.status <- c("blue","red","dark green")[batches]
plotMDS(gset, col = col.status, pch = 16)
```




## Design matrix

Now we are setting the theshold for the diferentially expressed genes
```{r}
PVALUE = 0.1
LOGFOLDCHANGE = 1
y = "Young"
o = "Old"
m = "MiddleAge"

EXPECTED_PROPORTION = 0.01

metadata_doc_name <- paste("samples_", my_id, ".csv", sep = "")
expression_doc_name <- paste("expression_", my_id, ".csv", sep = "")

assign_names <- function(c1, c2) {
  doc_name <- paste("DEG_list_", my_id, "_", c1, "_vs_", c2, ".csv", sep = "")
  dds_doc_name <- paste("dds_", my_id, "_", c1[1], c2[1], ".csv", sep = "")
  
  return (c(doc_name, dds_doc_name))
}

```

And let's save the expression and metadata files that are independent of the comparisons.
```{r saving_tables_1}
gene_table = featureData(gset)
symbol_column  <- gene_table$GB_ACC#ENSEMBL_ID
sequence_column <- gene_table$SEQUENCE
symbol_df <- data.frame(symbol = symbol_column, sequence= sequence_column)

# Combine the extracted "symbol" column with the row index from the AnnotatedDataFrame
ID_geneSymbol_df <- cbind(symbol_df, row.names = rownames(gene_table))



expression_df = merge(ex, ID_geneSymbol_df, by="row.names", all.x=TRUE)
write.table(expression_df, expression_doc_name, row.names=F, sep=",")

write.table(sample_data, metadata_doc_name, row.names=F, sep=",")

```
