---
title: "test"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(echo = TRUE, eval=FALSE)

```

# llll

```{r import, echo=FALSE}
library(illuminaHumanv4.db)
library(ggpubr)
library(GEOquery)
library(limma)
library(gplots)
library(dplyr)
library(tidyr)



```

## Importing the Data

```{r getting_experiment,  echo=FALSE}
my_id <- "GSE40551"
gse <- getGEO(my_id, GSEMatrix =TRUE)
```

Getting the samples from platform GPL570 from the experiment (in the bioproject)
```{r get_samples, echo=FALSE}
if (length(gse) > 1) idx <- grep("GPL16022", attr(gse, "names")) else idx <- 1
gset1 <- gse[[idx]]

gset = gset1

```


Seeing the metadata
```{r metadata}
pData(gset)
fData(gset)
fvarLabels(gset)
```
Seeing the headers of the metadata
```{r headers}
sample_data = pData(gset)
colnames(sample_data)

```

Here we need to manually select the column with the class that we are interesting and see the values
here the data seems to be in source_name_ch1

```{r see_unique_categories}
unique(sample_data$characteristics_ch1.1)
```

```{r see gender_categories}
unique(sample_data$`gender:ch1`)
```



In this case, source_name_ch1 has the class in a compose string.  I will separate them in age, sex (Female/Male) and one extra column age + age

```{r clean_up}
sample_data$age = sample_data$`treatment:ch2`
```

Getting the microarray ID's data (like the gene name)

```{r sample_metadata}
fvarLabels(gset) <- make.names(fvarLabels(gset))
features <- fData(gset)
features
```
Here, I will need a function to take a column and gives me a vector, 0 is it is young, 1 if it is middle age and 2 if it is old
```{r age_category_function}
map_age <- function(age_category) {
  
  age_vector <- ifelse(age_category == "before training",0, 1)
  #age_vector <- ifelse(age_category < 35, 0, ifelse(age_category >= 65, 2, 1))

  return (paste(age_vector, collapse = ""))
}

```

If we use it, we will get the mapped data

```{r }
# coding subtypes, 0 = YOUNG, 1 = MIDDLE AGE, 2 = OLD
gsms <- map_age(sample_data$age)
gsms #<- "111111111111000000000000"
sml <- strsplit(gsms, split="")[[1]]
```
And this is to put the labels when we don't have them, but in this case we do
```{r mapping}

gs <- factor(sml)
groups <- make.names(c("Obese", "Excersie"))
levels(gs) <- groups
gset$group <- gs
```
But since we did when we separated the title, we go with


## Check normalisation and scales used

Log2 scale transformation required

Check for normalisation by looking at boxplot of expression levels for each sample


```{r }
ex <- exprs(gset) # expression data, what form is this?
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
exprs(gset) <- log2(ex) } # i think this code is trying to determine whether the data is in log2 format yet
```

And here we see how are the samples before any selection
```{r boxplot1}

ord <- order(gs)  # order samples by group

palette <- palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste (my_id, "/", annotation(gset), sep ="")
boxplot(log2(ex[,ord]), boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")
```


```{r boxplot2}
boxplot(log2(ex),range=0,ylab="log2 intensity")
```


## Probe filtering

```{r }
# expression value distribution
par(mar=c(4,4,2,1))
title <- paste ("Normalized Expression Density", sep ="")
plotDensities(ex[,ord], group=gs, main=title, legend ="topright")
```

```{r}
# Probe filtering
medians <- rowMedians(ex)
man_threshold <- 4 

hist(medians, 150, col = "cornsilk1", freq = FALSE, 
main = "Histogram of the median intensities", 
border = "antiquewhite4",
xlab = "Median intensities")
```

```{r}
meta <- pData(gset)
meta$group
```

```{r}
no_of_samples <- table(meta$group)
sample_cutoff <- min(no_of_samples)

idx_man_threshold <- apply(Biobase::exprs(gset), 1, function(x){ sum(x > man_threshold) >= sample_cutoff})

table(idx_man_threshold)
```


```{r}
batches <- as.factor(gset$group)

col.status <- c("blue","red","dark green")[batches]
plotMDS(gset, col = col.status, pch = 16)
```




## Design matrix

Now we are setting the theshold for the diferentially expressed genes
```{r}
PVALUE = 0.1
LOGFOLDCHANGE = 1
y = "Young"
o = "Old"
m = "MiddleAge"

EXPECTED_PROPORTION = 0.01

metadata_doc_name <- paste("samples_", my_id, ".csv", sep = "")
expression_doc_name <- paste("expression_", my_id, ".csv", sep = "")

assign_names <- function(c1, c2) {
  doc_name <- paste("DEG_list_", my_id, "_", c1, "_vs_", c2, ".csv", sep = "")
  dds_doc_name <- paste("dds_", my_id, "_", c1[1], c2[1], ".csv", sep = "")
  
  return (c(doc_name, dds_doc_name))
}

```

And let's save the expression and metadata files that are independent of the comparisons.
```{r saving_tables_1}
gene_table = featureData(gset)
symbol_column  <- gene_table$ORF
symbol_df <- data.frame(symbol = symbol_column)
desired_columns <- c("GB_ACC", "SEQUENCE", "ENTREZ_GENE_ID", "RefSeq.Transcript.ID","REFSEQ", "Gene.Symbol", "Ensembl_ID", "SPOT_ID")

# Loop through each desired column
for (col_name in desired_columns) {
  # Check if the column exists in gene_table
  if (col_name %in% colnames(gene_table)) {
    symbol_column  <- gene_table[[col_name]]
    symbol_df[[col_name]] <- symbol_column
  } 
}
# Combine the extracted "symbol" column with the row index from the AnnotatedDataFrame
ID_geneSymbol_df <- cbind(symbol_df, row.names = rownames(gene_table))



expression_df = merge(ex, ID_geneSymbol_df, by="row.names", all.x=TRUE)
write.table(expression_df, expression_doc_name, row.names=F, sep=",")

write.table(sample_data, metadata_doc_name, row.names=F, sep=",")

```


And here we design the experiment with that expected proportion of genes, the gset and the comparisons, in this case we have 3 groups
```{r}
design <- model.matrix(~group + 0, gset)
colnames(design) <- levels(gs)
fit <- lmFit(gset, design)

# Set up contrasts of interest
cts <- paste(groups, c(tail(groups, -1), head(groups, 1)), sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
#fit2 <- eBayes(fit2, EXPECTED_PROPORTION)
fit2 <- eBayes(fit2)

# print contrasts order
colnames(fit2[["coefficients"]])
```

Here I am going to get one of the comparison of fit2 (see adove for the index), and specify the c's for the names 
```{r}
## Change this if needed
c1 = y
c2 = o
ncoef = 1
## and let this alone
docs_names = assign_names(c1,c2)
doc_name = docs_names[1]
dds_doc_name = docs_names[2]


```

and get the top genes
```{r}
#Create the top table for the comparison (coef) we want
top <- topTable(fit2, coef=ncoef, adjust.method = "fdr", sort.by="B", number = Inf)

#top_top <- top %>%
#  mutate(fdr = p.adjust(P.Value, "fdr"))
#baby_top <- topTable(fit2, coef=ncoef, adjust.method = "none", sort.by="B", number = Inf)
#top_top <- baby_top %>%
#  mutate(fdr = p.adjust(P.Value, "BH"))
#top_top <- top_top %>%
#  mutate(by = p.adjust(P.Value, "BY"))
#adjusted_all__p_values <- p.adjust(baby_top$p.value, method = "BH")

nrow(top)
```


Finally we are filtering by pvalue and LFC
```{r}
#Top Table of significant genes only
up_reg_epi <- subset(top, adj.P.Val < PVALUE, abs(logFC)>LOGFOLDCHANGE)
nrow(up_reg_epi)
```